---
title: "Pset 4"
author: "Andrea Moreno"
date: "2026-04-17"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(estimatr)
library(modelsummary)
card <- read_csv("data/card1995.csv")
```

1.2 
There are 3010 observations. The unit of observation is the individual, representing young men surveyed in the mid-1970s to study the relationship between education and labor market earnings. 

```{r}
#1.3
ggplot(card, aes(x = educ)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "Distribution of Years of Education",
       x = "Years of Education",
       y = "Frequency") +
  theme_minimal()
mean(card$educ, na.rm = TRUE)
```

1.3 
The mean is 13.26 years. In the historgram there are significant spikes at 12 and 16 years. These represent high school and four-year college completion, respectively.These spikes reflect milestones where the completion of a degree provides a specific credential for the labor market.

```{r}
#1.4
ggplot(card, aes(x = educ, y = lwage)) +
  geom_point(alpha = 0.2, color = "darkgrey") +  
  geom_smooth(method = "lm", color = "red", se = TRUE) +
  labs(title = "Naïve Relationship: Wages vs. Education",
       x = "Years of Education",
       y = "Log of Hourly Wage") +
  theme_minimal()
```
1.4
There is a clear positive correlation between years of education and log wages. The linear regression line slopes upward, suggesting that individuals with higher levels of schooling generally earn higher hourly wages.This naïve estimate may be biased if unobserved factors—such as underlying ability—are correlated with both schooling and earnings.

1.5
The OLS results might be misleading primarily because of omitted variable bias, specifically concerning innate ability. Individuals who possess high natural talent or motivation are more likely to pursue higher education and are also more likely to earn higher wages regardless of their schooling, which leads OLS to overstate the actual returns to education. Additionally, family background plays a significant role, students from wealthy or well connected families often have the financial support to stay in school longer and the social networks to land high-paying jobs, meaning the higher wages we observe might stem from their family's status rather than the degree itself.

```{r}
#2.1
mean(card$nearc4, na.rm = TRUE)
```
2.1
68.2% of respondents in this sample grew up near a 4-year college

```{r}
#2.2
first_stage <- lm_robust(educ ~ nearc4 + black + smsa + south + 
                         exper + expersq + married, 
                         data = card, se_type = "HC2")
modelsummary(first_stage, 
             statistic = c("std.error", "statistic"), 
             stars = TRUE)
```

2.2
The coefficient of 0.327 indicates that, holding all other factors constant, individuals who grew up near a four-year college completed approximately 0.33 more years of education on average compared to those who did not.Yes the positive sign makes perfect sense.Living near a college reduces the price of education by lowering travel costs and allowing students to live at home. When the cost of an activity decreases, we expect the quantity consumed to increase.

2.3
The t-statistic on nearc4 is 4.061, yielding a first-stage F-statistic of approximately 16.49. Since this exceeds the standard threshold of 10, the instrument is considered strong. Instrument strength is important because weak instruments lead to biased and imprecise 2SLS estimates potentially mirroring the same biases found in OLS.

2.4
a. Formally, the exclusion restriction requires that the instrument is uncorrelated with the error term in the structural equation.In the context of this study once we control for education and our other control variables, there must be no direct relationship between growing up near a college (nearc4) and log wages (lwage). Essentially, proximity to a college can only influence your paycheck by changing how many years you spent in school.
b. If people who grow up near colleges are more likely to stay in those often wealthier urban areas, their wages might be higher simply because of a higher cost of living or higher local pay scales, rather than their schooling.

```{r}
#2.5
card %>%
  group_by(nearc4) %>%
  summarize(
    mean_black = mean(black, na.rm = TRUE),
    mean_smsa  = mean(smsa, na.rm = TRUE),
    mean_south = mean(south, na.rm = TRUE),
    n = n()
  )
```
2.5
No, the groups are imbalanced. This imbalance reveals that growing up near a college is not randomly assigned.

```{r}
#3.1
reduced_form <- lm_robust(lwage ~ nearc4 + black + smsa + south + 
                          exper + expersq + married, 
                          data = card, se_type = "HC2")
modelsummary(reduced_form, 
             statistic = "std.error", 
             stars = TRUE)
```

3.1
The coefficient of 0.041 indicates that, after controlling for race, region, and experience, individuals who grew up near a four-year college earn approximately 4.1% higher hourly wages on average than those who did not.

3.2
.041/.327= .125
Wald = .125

3.3
The Wald estimate indicates that one additional year of education causes hourly wages to increase by approximately 12.5% for those individuals whose schooling decisions were influenced by their proximity to a college.

```{r}
ols <- lm_robust(lwage ~ educ + black + smsa + south + 
                 exper + expersq + married, 
                 data = card, se_type = "HC2")
modelsummary(ols, 
             statistic = "std.error", 
             stars = TRUE)
```

4.1 
Coefficient on education = 0.071

```{r}
#4.2
iv <- iv_robust(lwage ~ educ + black + smsa + south + 
                exper + expersq + married | 
                nearc4 + black + smsa + south + 
                exper + expersq + married, 
                data = card, se_type = "HC2")
modelsummary(iv, 
             statistic = "std.error", 
             stars = TRUE)
```
4.2
Coefficient on education = 0.125. Yes this matches my earlier Wald estimate. 

```{r}
#4.3
modelsummary(list("OLS" = ols, "2SLS" = iv),
             coef_map = c("educ" = "Years of Education"),
             gof_map = c("nobs", "r.squared"),
             stars = TRUE,
             title = "Comparison of OLS and 2SLS Estimates")
```

4.4
a. 2SLS is larger. 
b. The OLS estimate might be too low because of measurement error, where people misreport their years of schooling, which waters down the result and pushes it toward zero. Additionally, the compliers (the specific people influenced by living near a college) might be students from lower-income backgrounds who get a much larger than average boost in their career from that extra year of education.
c. The 2SLS estimate is more credible as a causal estimate. While it is less precise than OLS, it uses a natural experiment (proximity to college) to isolate the actual effect of schooling, whereas OLS likely suffers from ability bias or other unobserved factors that make it a simple correlation rather than a causal link.

```{r}
#4.5
ols_iq <- lm_robust(lwage ~ educ + IQ + black + smsa + south + 
                    exper + expersq + married, 
                    data = card, se_type = "HC2")
iv_iq <- iv_robust(lwage ~ educ + IQ + black + smsa + south + 
                   exper + expersq + married | 
                   nearc4 + IQ + black + smsa + south + 
                   exper + expersq + married, 
                   data = card, se_type = "HC2")
modelsummary(list("OLS (with IQ)" = ols_iq, "2SLS (with IQ)" = iv_iq),
             coef_map = c("educ" = "Years of Education", "IQ" = "IQ Score"),
             gof_map = c("nobs", "r.squared"),
             stars = TRUE)
```

4.5
a. The OLS has dropped after including IQ. This confirms that a small portion of what we thought was the return to education was actually just the effect of higher IQ scores. Since IQ and education are positively correlated, omitting IQ caused OLS to slightly overestimate the value of schooling.
b. The 2SLS estimate dropped from 0.125 to 0.109, but it is still much higher than the OLS estimate. Most importantly, the coefficient on IQ in the 2SLS model is 0.000 and is no longer statistically significant.
c. The comparison shows that IQ is not the main reason OLS and IV differ. Even after we directly control for IQ, the 2SLS estimate remains nearly double the OLS estimate. This suggests that the difference is likely driven by the fact that marginalized or low-income students who only go to college if one is nearby, get a much higher boost in wages from education than the average person does.

5.1
The compliers are people from lower-income families for whom the cost and convenience of a local college made the difference between attending or not. Unlike people who go to college regardless of distance, these individuals only stay in school because a four-year campus is close enough to eliminate the need for expensive room, board, or a long-distance move. Because they are often on the economic margin, that extra year of education provides a massive boost to their career trajectory, which explains why your LATE estimate is so much higher than the average.

5.2
Using the variation in education caused by growing up near a college, the 2SLS estimate indicates that one additional year of schooling causes a 10.9% increase in hourly wages specifically for those compliers who only pursued further education because they lived near a four year campus.

5.3
The LATE is likely higher than the ATE. This is because the compliers from lower income backgrounds who only stayed in school because a college was nearby—tend to have much higher returns to education than the average person. For these individuals, an extra year of schooling is a high leverage event that can move them from low-wage labor into professional careers, creating a huge percentage jump in earnings. In contrast, for the typical person an extra year of school adds some value but doesn't fundamentally transform their life trajectory to the same degree.
